Characterization of gaps in directly bonded Si compound optics

### 2014-03-18 AO_bondedSi.tex
lots of changes- add 2 figures, copious prose, minor tweaks

\subsection{Surface roughness or aberrations}
The second pitfall to consider is that the stiff non-flat Si surfaces will not conform upon contact leaving the complement of the aberrations as an air gap. Similarly, surface roughness will also limit the bonding strength  \cite{JJAP.37.4197,2001JOptA...3...85G}.  The details of this aspect of the bonding process are outside the scope of this paper.

\section{Silicon-gap-silicon Fabry-P\`{e}rot theory}
\label{secTheory}

Our measurement technique exploits the large Fresnel reflection \cite{2001opt4.book.....H} at the Si-Air interface.  The refractive index of silicon at room temperature ranges from 3.55$-$3.45 from $\lambda = $ 1150$-$2500 nm \cite{2006SPIE.6273E..77F}, one of the largest refractive indices of any conventional material.  The Fresnel reflection is accordingly about 30\% per surface.  Popular optical materials with refractive indices in the range 1.5$-$1.7 with Fresnel reflection 4$-$7\% per surface will suffer less than Si from the effects described here.

\section{Imbedding gaps of known sizes}
We evaluated our metrology method by imbedding gaps of known sizes in the interface.  We bored divots in silicon with inductively coupled plasma etching.  The tool was a Unaxis.  We used two different gases with low and high etch rates.  CHF$_3$ on Si exhibited about 0.3 nm/second etch rate, SF$_6$ exhibited about 1 micron/minute.  The area exposed to the process gases was about 1 inch diameter circles, or 15 mm $\times$ 20 mm rectangles.  We protected the unexposed areas with thick photoresist.  We achieved etch depths of 20$\pm$5 nm and 4000$\pm$100 nm, as measured with Wyko 2D noncontact white light microscopy for the small depth, and Dektak stylus profilometry for the large depth.  For measurements taken with the Wyko, uncertainties were constructed by inspecting the histogram of topology.  The large-scale distortions were removed from the topology by masking and flattening to ensure that the uncertainties accurately reflect the distribution of measured heights.  The stylus profiliometry measured values were simply assigned an uncertainty of 5\%, which was consistent with our intuition and experience.

@@

\section{Transmission measurement theory}
The Si-air\footnote{We imprecisely refer to the gap as an air gap, though in principle the gap has an unknown mixture of gases for example water vapor, or other vapors that were outgassed from reactions catalyzed by hydrocarbons.} interfaces of the in the bond gap make a Fabry-P\`erot etalon\cite{2007fuph.book.....S}.  The transmission through a Fabry-Perot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  We assume transmission is at normal incidence, the index of the gap is 1.0, and the material is at room temperature.  We assume Si is nonabsorbing longward of $\lambda$ = 1150 nm.  We compute the reflectivity for an Si-air interface, which is a modest function of wavelength through the refractive index of Silicon, $n(\lambda, T)$ \cite{2006SPIE.6273E..77F}.  The computation is simply Fresnel's law at normal incidence:

@@

\section{Substrate preparation}
It is important to clean the wafers to minimize the particle density, since particles cause gaps.  The substrates were silicon wafers.  The surface roughness was typically about 2 nm, as measured with a Wyko 2D noncontact microscope.  The substrate thicknesses ranged from 0.5 to 3.3 mm per substrate.  We classify the 3.3 mm thick substrates as ``pucks'' and the $\sim$1.0 mm thick substrates as ``wafers''.  The distinction between the two classes of substrate thickness is based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is much slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.  We measured the large scale surface flatness of the pucks.  The Fisba2 interferometer has a 2 inch diameter beam at $\lambda=$632.8 nm.  We found a typical peak to valley surface flatness of 2 waves over the central 2 inch diameter.

@@

Our sensitive measurement technique introduces a few challenges at the parts per thousand level baseline level.  Specifically, we have been ambivalent about whether or not to use the double-beam capability of the Cary5000.  We have shown that this mode reduces by a factor of 10, the random flicker attributable to lamp noise.  In practice, the double beam mode can cause some systematic problems near the portions of wavelengths near lamp and filter changes.  So of these issues might be related to polarization differences, and a way to mitigate these techniques is to follow the advice of the vendor and employ a depolarizer in the beam path.  A depolarizer strategy is critical for applications in which a non-normal incidence angle is desired.  A half-hearted attempt to measure non-normal incidence without the polarization accessory resulted in failure.  In the next section we describe our interest in non-normal incidence.

\subsection{Non-normal incidence angles}
For many applications, normal incidence performance is the quantity of interest.  The technique described here outlines an approach for measuring the gap size of bonded Si wafers at normal incidence.  In principle, from there the performance at non-normal incidence can be worked out either by simulation or calculation.  Here we briefly motivate the challenge of experiment and simulation at non-normal incidence.

Our final application that motivated this project is to make diffraction gratings on the hypotenuse of a silicon prism.  The diffraction occurs inside the silicon material and is dispersed and reflected back along its path.  This configuration is called an immersion grating.  Our group has made monolithic immersion gratings for over a decade, but challenges in processing large (30 mm thick) Si wafers motivated an approach of patterning and processing thin (1-10 mm thick) Si wafers, and then bonding them to shaped and polished prisms.  This technique has been pioneered at SRON.

The consequence of our design is a non-normal incidence interface in double pass.  In this case, the framework developed here fails to predict the spectrum of the interface loss attributable to a finite albeit small (few nm) gap.  Specifically, at large incidence angles complete internal reflection and evanescent waves will matter.  Polarization effects are important.

>>>>

\subsection{Surface roughness or large-scale aberrations}
The second pitfall to consider is that the stiff non-flat Si surfaces will not conform upon contact leaving the complement of the aberrations as an air gap. This issue is much less of a problem for wafers compared to our unconventional thick ($\sim$ 30 mm) Si applications.  Similarly, surface roughness will also limit the bonding strength  \cite{JJAP.37.4197,2001JOptA...3...85G}.  The details of this aspect of the bonding process are outside the scope of this paper.  In short, our thick substrates are optically polished to much less than $\lambda = 632$ nm peak to valley over the 3 to 4 inch diameter clear aperture.  We expect such well-polished substrates to have no deleterious impact on the bonding process.

\section{Imbedding gaps of known sizes}
We evaluated our metrology method by imbedding gaps of known sizes in the interface.  We bored divots in silicon with inductively coupled plasma etching.  The tool was a Unaxis.  We used two different gases with low and high etch rates.  CHF$_3$ on Si exhibited about 0.3 nm/second etch rate, SF$_6$ exhibited about 1 micron/minute.  The area exposed to the process gases was about 1 inch diameter circles, or 15 mm $\times$ 20 mm rectangles.  We protected the unexposed areas with thick photoresist.  We achieved etch depths of 20$\pm$5 nm and 4000$\pm$100 nm, as measured with Veeco NT9100 Optical Profiler for the small depth, and Dektak stylus profilometry for the large depth.  For measurements taken with the Veeco, uncertainties were constructed by inspecting the histogram of topology.  The large-scale distortions were removed from the topology by masking and flattening to ensure that the uncertainties accurately reflect the distribution of measured heights.  The stylus profiliometry measured values were simply assigned an uncertainty of 5\%, which was consistent with our intuition and experience.

@@

\section{Transmission measurement theory}
\label{secTheory}

% Figures (shared wavelength axis)
%	1) Refractive index as a function of wavelength at room temp
%	2) Fresnel transmission as a function of wavelength (normal incidence)
%	3) Coefficient of finesse as a function of wavelength
%	4) Example transmission through VG03 thick gap

Our measurement technique exploits the large Fresnel reflection \cite{2001opt4.book.....H} at the Si-Air interface.  The refractive index of silicon at room temperature ranges from 3.55$-$3.45 from $\lambda = $ 1150$-$2500 nm \cite{2006SPIE.6273E..77F}, one of the largest refractive indices of any conventional material.  The Fresnel reflection is accordingly about 30\% per surface.  For comparison, conventional optical materials with refractive indices in the range 1.5$-$1.7 have Fresnel reflection of merely 4$-$7\% per surface.  Figure X shows the refractive index as a function of wavelength for room temperature, which we computed using the coefficients from \cite{2006SPIE.6273E..77F}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.95\columnwidth]{figs/SiIndexAOmgsFinesseFig.pdf}}
\caption{Plot of refractive index, $n$, and coefficient of finesse, $F$, as a function of wavelength, $\lambda$.\label{fig} We computed the temperature dependent refractive index from the tabulations of \cite{2006SPIE.6273E..77F}, and extrapolated past their upper temperature bound to estimate values near 373 K.  The coefficient of finesse is defined in Equation \ref{eq:FabPerot}.  The minuscule temperature dependence is largely negligible, and for the rest of the paper we assume room temperature refractive index.}
\end{figure}

The Si-air\footnote{We imprecisely refer to the gap as an air gap, though in principle the gap has an unknown mixture of gases for example water vapor, or other vapors that were outgassed from reactions catalyzed by hydrocarbons.} interfaces of the in the bond gap make a Fabry-P\`erot etalon\cite{2007fuph.book.....S}.  The transmission through a Fabry-Perot depends on the wavelength of light, the reflectivity of the etalon sidewalls, and the size of the gap.  We assume transmission is at normal incidence, the index of the gap is 1.0, and the material is at room temperature.  We assume Si is nonabsorbing longward of $\lambda$ = 1150 nm.  We compute the reflectivity for an Si-air interface, which is a modest function of wavelength through the refractive index of Silicon, $n(\lambda, T)$ \cite{2006SPIE.6273E..77F}.  The computation is simply Fresnel's law at normal incidence:

@@

\section{Substrate preparation}
It is important to clean the wafers to minimize the particle density, since particles cause gaps.  The substrates were silicon wafers.  The surface roughness was typically about 2 nm, as measured with a Veeco NT9100 Optical Profiler .  The substrate thicknesses ranged from 0.5 to 3.3 mm per substrate.  We classify the 3.3 mm thick substrates as ``pucks'' and the $\sim$1.0 mm thick substrates as ``wafers''.  The distinction between the two classes of substrate thickness is based on the anticipation that thin substrates will conform more easily to their bonding partner substrates.  There is also evidence that bond front propagation is much slower in thick pucks \cite{2007ApOpt..46.6793H} than in wafers.  We measured the large scale surface flatness of the pucks.  The Fisba2 interferometer has a 2 inch diameter beam at $\lambda=$632.8 nm.  We found a typical peak to valley surface flatness of 2 waves over the central 2 inch diameter.

@@

\subsection{Non-normal incidence angles}
For the vast majority of applications, normal incidence performance is the quantity of interest.  The technique laid out in this paper outlines an approach for measuring the gap size of bonded Si wafers at normal incidence.  In principle, from there the performance at non-normal incidence can be worked out either by simulation or calculation.  Here we briefly motivate the challenge of experiment and simulation at non-normal incidence.

\begin{figure}[htbp]
\centerline{\includegraphics[width=0.95\columnwidth]{figs/Sibondedprismcartoon.pdf}}
\caption{Schematic of a bonded Si immersion grating\label{figSiPrism}.  }
\end{figure}

Our final application that motivated this project is to make diffraction gratings on the hypotenuse of a silicon prism.  The diffraction occurs inside the silicon material and is dispersed and reflected back along its path.  This configuration is called an immersion grating.  Figure \ref{figSiPrism} shows a schematic of the geometry of our bonded Si immersion grating.  Our group has made monolithic immersion gratings for over a decade, but challenges in processing large (30 mm thick) Si wafers motivated an approach of patterning and processing thin (1-10 mm thick) Si wafers, and then bonding them to shaped and polished prisms.  This technique has been pioneered at SRON \cite{2012SPIE.8450E..2TV}.  In Figure \ref{figSiPrism} we show an example with a 10 mm thick Si puck bonded to a 25 mm thick, precision cut, shaped, and polished Si prism.  The group at SRON has patterned Si immersion echelle gratings on commercial thin  ($< 1$ mm) Si wafers.  Thin wafers have the drawback that they are conformal, so total-thickness-variations (TTV) directly translate into surface optical aberrations\cite{2012SPIE.8450E..2TV}.

The consequence of our immersion grating design is a non-normal incidence interface in double pass.  In this case, the Fabry-P\'{e}rot framework developed here fails to predict the spectrum of the interface loss attributable to a finite albeit small (few nm) gap.  Specifically, complete internal reflection and evanescent waves will matter at large incidence angles, and the incoherent matrix technique is no longer accurate.  Polarization effects are important.  We do not have a mechanism for evaluating the bond performance until the final device is completed and its wavelength-dependent transmission is measured.  We can detect the performance degradation on final devices in the following nuanced way.  We will measure the efficiency of the bonded Si immersion grating as a function of both wavelength and angle.  Our group built a custom tool to make these measurements- the Custom Robotic Order, Wavelength, and Blaze Angle Recorder (CROWBAR) \cite{2012SPIE.8450E..2SG}.  Figure 3 in \cite{2012SPIE.8450E..2SG} shows the directly measured spectrum of a monolithic immersion grating.  Notably, the spectrum in Figure 3 in \cite{2012SPIE.8450E..2SG}  shows dozens of blaze peaks where the efficiency should be maximized.  A perfectly contacted Si$-$Si direct bonded optic would have identical performance to a monolithic counterpart.  Bonded immersion gratings with finite gap sizes will superimpose a wavelength dependent transmission loss on top of the blaze peaks.  In this way, the presence or absence of deleterious gaps can be identified in final immersion gratings.  One drawback of this approach is that gaps can only be identified after the costly process of producing the final optical device.  Large gaps identified this late in the game could render the costly optic useless, both in terms of efficiency and wavefront performance.  Our group is exploring the effectiveness of high-temperature annealing to eliminate IR bubbles in completed devices.
